Natural selection in the Health and Retirement Study

### 2023-09-16 hrs-selection.qmd
Move pics to appendix

---
title: "Natural selection in the Health and Retirement Study"
author: "David Hugh-Jones"
abstract: |
  I investigate natural selection on polygenic scores
  in the contemporary US, using the Health and Retirement Study. Results
  support the economic theory of fertility as an explanation for
  natural selection: among both white and black respondents,
  scores which correlate negatively (positively) with education are
  selected for (against). However, selection effects 
  are not larger among people with less education. 
  For whites, selection effects are smaller among married people. 
date: "September 2023"
format: 
  pdf:
    geometry:
    - top=2in
    - left=1.5in
    - right=1.5in
    - heightrounded
editor: visual
editor_options: 
  chunk_output_type: console
knitr:
  opts_chunk: 
    echo: false
bibliography: bibliography-hrs.bib
---

@@

# TODO: siblings E5669 "how many have died"
#   rands$white |> select(matches("r\\d+livsib")) -> tmp
#   apply(tmp, 1, max)

# TODO: use survey package
# here's some useful code https://github.com/ajdamico/asdfree/blob/master/27-hrs.Rmd

@@

# TODO: age at first birth. RAND family data provides h*ageokid
# TODO: RAND family data has r*sbdied for how many siblings have died

@@

library(haven)
library(forcats)
# library(Hmisc)
library(dplyr, warn.conflicts = FALSE)
library(stringr)
library(purrr)
library(ggplot2)
library(tidyr)
library(santoku, warn.conflicts = FALSE) # masks tidyr::chop

@@

titles <- list(white = "Whites (non-Hispanic)", black = "Blacks")

@@

eths <- c("white", "black")
pgens <- list(white = pgen4e, black = pgen4a)

@@

@fig-fertility shows effect sizes of PGS correlations with realised fertility among each ethnic group. Standard errors are large because of the relatively low sample sizes, and only `r n_sig_bonf` score is significant at Bonferroni-corrected p \< 0.05/`r n_tests`. But I am most concerned with looking at patterns across scores rather than judging the significance of individual scores. (Also, because all scores are differently weighted sums of the same set of alleles, if we can reject the null for any score we should probably reject it for all of them; in other words, effect sizes are more interesting than p values here.)

@@

for (eth in eths) {
  ggp <- ggplot(tidy_pgs_joined[[eth]], aes(estimate.edyrs, estimate.fert)) +
          geom_hline(yintercept = 0, color = "grey80") +
          geom_vline(xintercept = 0, color = "grey80") +
          geom_point() +
          geom_smooth(method = "lm", formula = y ~ x, se = FALSE, 
                      linetype = "dashed", colour = "red", linewidth = 0.5) +
          scale_y_continuous(breaks = seq(-0.35, 0.15, 0.05),
                             labels = scales::label_number(0.01)) +
          coord_cartesian(xlim = c(-.3, .75)) +
          theme_light() + 
          theme(panel.grid = element_blank()) +
          labs(
            title = titles[[eth]],
            x = "Effect on years of education",
            y = "Effect on realized fertility"
          )
  print(ggp)
}

@@

I can also examine natural selection in the previous generation, by regressing PGS on respondents' number of siblings (including dead siblings) in 2010. I reweight respondents by the reciprocal of their number of siblings, to account for parents of many siblings being more likely to be a parent of a respondent. Parents of no siblings cannot be included, so effect sizes are not comparable across the generations. @fig-siblings plots effects on number of siblings versus effects on years of education. The correlation is negative and significant for whites (correlation `r cors_sibs_edu$white`, bootstrapped 95% C.I. `r tails_boot_sibs$white[1]` to `r tails_boot_sibs$white[2]`). It is positive for blacks (correlation `r cors_sibs_edu$black`), although this is driven by two outlier PGS, which are scores for Alzheimer's with all SNPs included; without these, the correlation is negative.

@@

```{r}
#| label: stat-x-edu
#| cache: true

@@

nobs <- map_dbl(eths, 
                \(x) nobs(mod_pgs_x_agefb[[x]][[1]])) |> 
  setNames(eths)

tidy_pgs_x_ed |> 
  list_rbind(names_to = "ethnicity") |> 
  mutate(
    Ethnicity = paste0(unlist(titles[ethnicity]), 
                       " (N = ", nobs[ethnicity], ")"),
    Education = str_remove(edu, "edu")
  ) |> 
  ggplot(aes(Education, abs(estimate), color = Education, fill = Education)) +
    geom_point(position = position_jitter(width = 0.05), alpha = 0.8) +
    geom_violin(alpha = 0.4, draw_quantiles = 1:3/4) +
    facet_wrap(vars(Ethnicity), scales = "free_y") +
    labs(
      x = "Years of education",
      y = "Absolute effect size"
    ) +
    theme_minimal() +
    theme(
      legend.position = "none"
    )
```

@@

tidy_pgs_x_inc |> 
  list_rbind(names_to = "ethnicity") |> 
  mutate(
    Ethnicity = unlist(titles[ethnicity]),
    Income = fct_relevel(income_med, c("Low", "High"))
  ) |> 
  ggplot(aes(Income, abs(estimate), color = Income, fill = Income)) +
    geom_point(position = position_jitter(width = 0.05), alpha = 0.8) +
    geom_violin(alpha = 0.4, draw_quantiles = 1:3/4) +
    facet_wrap(vars(Ethnicity), scales = "free_y") +
    labs(
      x = "Income",
      y = "Absolute effect size"
    ) +
    theme_minimal() +
    theme(
      legend.position = "none"
    )
```

@@

tidy_pgs_x_mar |> 
    list_rbind(names_to = "ethnicity") |> 
    mutate(
        Ethnicity = unlist(titles[ethnicity]),
        married = fct_relevel(married, "Other", "Married")
    ) |> 
    ggplot(aes(married, abs(estimate), color = married, fill = married)) +
      geom_point(position = position_jitter(width = 0.05), alpha = 0.8) +
      geom_violin(alpha = 0.4, draw_quantiles = 1:3/4) +
      facet_wrap(vars(Ethnicity), scales = "free_y") +
      labs(
          x = "Married",
          y = "Absolute effect size"
      ) +
      theme_minimal() +
      theme(
          legend.position = "none"
      )

@@

tidy_pgs_x_agefb |> 
    list_rbind(names_to = "ethnicity") |> 
    mutate(
        Ethnicity = unlist(titles[ethnicity]),
        agefb = fct_relevel(agefb, "Low", "High")
    ) |> 
    ggplot(aes(agefb, abs(estimate), color = agefb, fill = agefb)) +
      geom_point(position = position_jitter(width = 0.05), alpha = 0.8) +
      geom_violin(alpha = 0.4, draw_quantiles = 1:3/4) +
      facet_wrap(vars(Ethnicity), scales = "free_y") +
      labs(
          x = "Age at first birth",
          y = "Absolute effect size"
      ) +
      theme_minimal() +
      theme(
          legend.position = "none"
      )

@@

@@

>>>>

---
title: "Natural selection in the Health and Retirement Study"
author: "David Hugh-Jones"
abstract: |
  I investigate natural selection on polygenic scores
  in the contemporary US, using the Health and Retirement Study. Results
  support the economic theory of fertility as an explanation for
  natural selection: among both white and black respondents,
  scores which correlate negatively (positively) with education are
  selected for (against). However, selection effects 
  are not larger among people with less education. 
  For whites, selection effects are smaller among married people. 
date: "September 2023"
format: 
  pdf:
    include-in-header:
      - text: |
          \usepackage{placeins}
    geometry:
    - top=2in
    - left=1.5in
    - right=1.5in
    - heightrounded
editor: visual
editor_options: 
  chunk_output_type: console
knitr:
  opts_chunk: 
    echo: false
bibliography: bibliography-hrs.bib
---

@@

# TODO: use survey package
# here's some useful code https://github.com/ajdamico/asdfree/blob/master/27-hrs.Rmd
# TODO (related) move away from lists and towards a single datasource with an
# "ethnicity" column; results from broom::tidy can still be in lists?

@@

# TODO: think about your bootstraps, what's the test... given that 
# the scores are not independent, does it really make sense to aggregate
# a general size comparison? Or am I secretly just reproducing my own
# crappy standard errors?

@@

library(haven)
library(forcats)
library(dplyr, warn.conflicts = FALSE)
library(stringr)
library(purrr)
library(ggplot2)
library(tidyr)
library(santoku, warn.conflicts = FALSE) # masks tidyr::chop

@@

eths <- c("white", "black")

titles <- c(white = "White", black = "Black")

@@

pgens <- list(white = pgen4e, black = pgen4a)

@@

Appendix @fig-fertility shows effect sizes of PGS correlations with realised fertility among each ethnic group. Standard errors are large because of the relatively low sample sizes, and only `r n_sig_bonf` score is significant at Bonferroni-corrected p \< 0.05/`r n_tests`. But I am most concerned with looking at patterns across scores rather than judging the significance of individual scores.

@@

for (eth in eths) {
  ggp <- ggplot(tidy_pgs_joined[[eth]], aes(estimate.edyrs, estimate.fert)) +
          geom_hline(yintercept = 0, color = "grey80") +
          geom_vline(xintercept = 0, color = "grey80") +
          geom_point() +
          geom_smooth(method = "lm", formula = y ~ x, se = FALSE, 
                      linetype = "dashed", colour = "red", linewidth = 0.5) +
          scale_y_continuous(breaks = seq(-0.35, 0.15, 0.05),
                             labels = scales::label_number(0.01)) +
          coord_cartesian(xlim = c(-.3, .75)) +
          theme_light() + 
          theme(panel.grid = element_blank()) +
          labs(
            title = titles[eth],
            x = "Effect on years of education",
            y = "Effect on realized fertility"
          )
  print(ggp)
}

@@

I can also examine natural selection in the previous generation, by regressing PGS on respondents' number of siblings (including dead siblings) in 2010. I reweight respondents by the reciprocal of their number of siblings, to account for parents of many siblings being more likely to be a parent of a respondent. Parents of no siblings cannot be included, so effect sizes are not comparable across the generations. Appendix @fig-siblings plots effects on number of siblings versus effects on years of education. The correlation is negative and significant for whites (correlation `r cors_sibs_edu$white`, bootstrapped 95% C.I. `r tails_boot_sibs$white[1]` to `r tails_boot_sibs$white[2]`). It is positive for blacks (correlation `r cors_sibs_edu$black`), although this is driven by two outlier PGS, which are scores for Alzheimer's with all SNPs included; without these, the correlation is negative.

@@

```{r}
#| label: stat-x-edu-bootstraps
#| cache: true

@@

tidy_pgs_x_ed |> 
  list_rbind(names_to = "ethnicity") |> 
  mutate(
    Ethnicity = paste(titles[ethnicity], calc_n(mod_pgs_x_ed, ethnicity)),
    Education = str_remove(edu, "edu")
  ) |> 
  ggplot(aes(Education, abs(estimate), color = Education, fill = Education)) +
    geom_point(position = position_jitter(width = 0.05), alpha = 0.8) +
    geom_violin(alpha = 0.4, draw_quantiles = 1:3/4) +
    facet_wrap(vars(Ethnicity), scales = "free_y") +
    labs(
      x = "Years of education",
      y = "Absolute effect size"
    ) +
    theme_minimal() +
    theme(
      legend.position = "none"
    )
```

@@

tidy_pgs_x_inc |> 
  list_rbind(names_to = "ethnicity") |> 
  mutate(
    Ethnicity = paste(titles[ethnicity], calc_n(mod_pgs_x_inc, ethnicity)),
    Income = fct_relevel(income_med, c("Low", "High"))
  ) |> 
  ggplot(aes(Income, abs(estimate), color = Income, fill = Income)) +
    geom_point(position = position_jitter(width = 0.05), alpha = 0.8) +
    geom_violin(alpha = 0.4, draw_quantiles = 1:3/4) +
    facet_wrap(vars(Ethnicity), scales = "free_y") +
    labs(
      x = "Income",
      y = "Absolute effect size"
    ) +
    theme_minimal() +
    theme(
      legend.position = "none"
    )
```

@@

tidy_pgs_x_mar |> 
    list_rbind(names_to = "ethnicity") |> 
    mutate(
      Ethnicity = paste(titles[ethnicity], calc_n(mod_pgs_x_mar, ethnicity)),
      married = fct_relevel(married, "Other", "Married")
    ) |> 
    ggplot(aes(married, abs(estimate), color = married, fill = married)) +
      geom_point(position = position_jitter(width = 0.05), alpha = 0.8) +
      geom_violin(alpha = 0.4, draw_quantiles = 1:3/4) +
      facet_wrap(vars(Ethnicity), scales = "free_y") +
      labs(
          x = "Married",
          y = "Absolute effect size"
      ) +
      theme_minimal() +
      theme(
          legend.position = "none"
      )

@@

tidy_pgs_x_agefb |> 
    list_rbind(names_to = "ethnicity") |> 
    mutate(
     Ethnicity = paste(titles[ethnicity], calc_n(mod_pgs_x_agefb, ethnicity)),
      agefb = fct_relevel(agefb, "Low", "High")
    ) |> 
    ggplot(aes(agefb, abs(estimate), color = agefb, fill = agefb)) +
      geom_point(position = position_jitter(width = 0.05), alpha = 0.8) +
      geom_violin(alpha = 0.4, draw_quantiles = 1:3/4) +
      facet_wrap(vars(Ethnicity), scales = "free_y") +
      labs(
          x = "Age at first birth",
          y = "Absolute effect size"
      ) +
      theme_minimal() +
      theme(
          legend.position = "none"
      )

@@

\FloatBarrier

@@

## Figures
